---
title: "ASSIGNMENT 2"
subtitle: "DSCI 6607 -- Programmatic Data Analysis Using Python and R  \nName: Sahil
  Khan  \nStudent ID: 202482066   \nSubmission Date: 26-Oct-2024 \n"
output:
  pdf_document:
    latex_engine: xelatex
---

## Question 1  
Recall the bisection method, we leaned in assignment 1. The bisection method can be generalized to deal
with the case f(xl)f(xr) > 0 (i.e., the two end points do not have opposite signs), by broadening the bracket.
That is, we reduce xl and/or increase xr, and try again. A reasonable choice for broadening the bracket is to
double the width of the interval [xl, xr], that is
$$ {\text m} \leftarrow {(x_l + x_r)}/2, $$
$${\text w} \leftarrow {{x_r - x_l}}, $$
$${x_l} \leftarrow {{m - w}}, $$
$${x_r} \leftarrow {{m + w}}, $$
\begin{itemize}
\item[a)] Incorporate bracket broadening into the bisection method 
\end{itemize}
```{r}
bisectionBroadening <- function(f, xl, xr, tol = 1e-6, max_iter = 50) {
    iter <- 0
    while (iter < max_iter) {
        m <- (xl + xr) / 2
        w <- xr - xl
        if (f(xl) * f(xr) <= 0) {
            break
        }
        xl <- m - w
        xr <- m + w
        iter <- iter + 1
    }
    while (abs(xr - xl) > tol && iter < max_iter) {
        m <- (xl + xr) / 2
        if (f(xl) * f(m) < 0) {
            xr <- m
        } else {
            xl <- m
        }
        iter <- iter + 1
    }
    return(m)
}
```
\begin{itemize}
\item[b)] Finding the root of
$$ f(x) = (x − 1)^3 − 2x^2 + 10 − sin(x) $$ 
\end{itemize}
```{r}
f <- function(x) (x - 1)^3 - 2 * x^2 + 10 - sin(x)
root <- bisectionBroadening(f, 1, 2)
print(root)
```
***

## Question 2
We plan to test the equality for the means of two samples in Python.
\begin{itemize}
\item[1.] Let x and y be two samples of sizes n1 and n2, respectively. To test $$ H0 : μx = μy $$ vs $$ H1 : μx \neq μy, $$ we compute the test statistic
$$ t = \frac{\bar{x} - \bar{y}}{\sqrt{s_p^2(\frac{1}{n_1}+\frac{1}{n_2})}} $$
where $$ \bar{x}, \bar{y} $$ denote the means of x and y samples and
$$ s_p^2 = \frac{(n_1-1)S_1^2 +(n_2-1)S_2^2 }{n_1 + n_2 -1} $$
where
$$ s_1^2 = \frac{1}{n_1-1} \sum_{i=1}^{n_1}(x_i - \bar{x})^2 $$
$$ s_2^2 = \frac{1}{n_2-1} \sum_{i=1}^{n_2}(y_i - \bar{y})^2 $$
\end{itemize}

\begin{itemize}
\item[2.] Python function which takes x and y as two lists and returns the observed test statistic.
\end{itemize}
```{python}
import numpy as np

def tStatistic(x, y):

  n1 = len(x)
  n2 = len(y)
  xBar = np.mean(x)
  yBar = np.mean(y)
  s1 = np.var(x, ddof=1)
  s2 = np.var(y, ddof=1)
  sp = ((n1-1)*s1 + (n2-1)*s2) / (n1 + n2 - 2)
  t = (xBar - yBar) / np.sqrt(sp * (1/n1 + 1/n2))

  return t
```

\begin{itemize}
\item[3.] Generate 50 observations form normal distribution with mean =1 and standard deviation=2 and assign the data to list x. Generate 57 observations from uniform distribution between -2 and 2 and assign data to list y.
\end{itemize}
```{python}
# Generating sample X from Normal Distribution
x = np.random.normal(loc=1, scale=2, size=50)

# Generating sample Y from Uniform Distribution
y = np.random.uniform(low=-2, high=2, size=57)
```

\begin{itemize}
\item[4.] Applying our function and computing the test statistic for the two samples from part 3.
\end{itemize}
```{python}

t = tStatistic(x, y)

print("The observed t-statistic is:", t)
```
***

## Question 3
Consider the python list
$$ x = [3, 8, 13, 18, 108, 25, 23, 17, 203, 11, 23] $$
Write a python function where takes the list and uses only list comprehension and returns the odd values
smaller than 23.
```{python}
def oddValuesUnder23(lst):
  return [x for x in lst if x % 2 != 0 and x < 23]

x = [3, 8, 13, 18, 108, 25, 23, 17, 203, 11, 23]

print("Odd values smaller than 23:", oddValuesUnder23(x))
```
***

## Question 4
\begin{itemize}
\item[1.] Find the distribution of statistic $$ \sum_{i=1}^{10}X_i $$
\end{itemize}
$$ \text{Given that} X_i \sim N(\mu, \sigma^2) \text{ for i=1,2,...,10 where } \mu \text{ and } \sigma^2 \text{ are the mean and variance of the population,}$$ 
$$ \text{we want to find the distribution of statistics} \sum_{i=1}^{10}X_i $$
Let: $$ S = \sum_{i=1}^{10}X_i $$
Since each Xi is normally distributed, and the sum of independent normal random variables is also normally distributed, S will be normally distributed.

Mean of the Sum: The mean of S, E(S) is the sum of the means of Xi.
Variance of the Sum: The variance of S, Var(S), is the sum of the variances of Xi.
$$ E(S) = E(\sum_{i=1}^{10}X_i) = \sum_{i=1}^{10}E(X_i) = \sum_{i=1}^{10}\mu = 10\mu $$
$$ Var(S) = Var(\sum_{i=1}^{10}X_i) = \sum_{i=1}^{10}Var(X_i) = \sum_{i=1}^{10}\sigma^2 = 10\sigma^2 $$
Therefore,
$$ S = \sum_{i=1}^{10}X_i \sim N(10\mu,10\sigma^2) $$
\begin{itemize}
\item[2.] We would like to test the finding of part (1) numerically. To do that first generate a sample of size 10 from the Normal distribution with parameters $$ \mu = 23 \text{ and } \sigma^2 = 3.6$$ and then compute the sum of the generated observations.
\end{itemize}
```{python}
def genNormalDist(sims, size=10, mu=23, sigSq=3.6):

	sampleSums = []

	for i in range(sims):
	  # Generating a sample of size=10 from Normal Distribution
		sample = np.random.normal(loc=mu, scale=np.sqrt(sigSq), size=size)

		sampleSums.append(np.sum(sample))

	return sampleSums

print("Sample Sum:", genNormalDist(sims=1))
```

\begin{itemize}
\item[3.] Use python and simulate 10000 times part (2) and compute the sum of the generated samples of size 10.
\end{itemize}
```{python}
simulations = 10000
sampleSumList = genNormalDist(sims=simulations)
print("Sample Sums after 10000 simulations (first 10 obs):")
print("\n".join(str(sum_value) for sum_value in sampleSumList[0:10]))
```
\begin{itemize}
\item[4.] Plot the histogram of the 10000 observed statistics from part (3). Then show the density curve of the theoretical distribution you found in part (1) on the histogram.
\end{itemize}
```{python}
import matplotlib.pyplot as plt
from scipy.stats import norm

mu=23
sigSq=3.6 

plt.hist(sampleSumList, bins=50, density=True, alpha=0.5, label='Simulated')

# Plotting the density curve of the Theoretical Distribution
x = np.linspace(np.min(sampleSumList), np.max(sampleSumList), 100)
plt.plot(x, norm.pdf(x, loc=10*mu, scale=np.sqrt(10*sigSq)), color='red', label='Theoretical')

plt.xlabel('Sum of 10 Observations')
plt.ylabel('Probability Density')
plt.legend()
plt.title('Histogram of Simulated and Theoretical Distribution')
plt.show()
```

***
## Question 5
The continuous random variable X has the following probability density function (pdf), for some positive
constant c,
$$ f(x) = \frac{3}{(1+x)^3},\ \ 0\leq x \leq c. $$
\begin{itemize}
\item[a.] Find c which makes f a legitimate pdf?
\end{itemize}
To find the constant c that makes the fxn f(x) a legitimate pdf over the interval [0,c], we need to make sure that the total area under the pdf curve from 0 to c equals 1.
$$ \int_{0}^{c} f(x)dx = 1 $$
$$ \int_{0}^{c} \frac{3}{(1+x)^3}dx = 1 $$
$$ \int \frac{3}{(1+x)^3}dx=-\frac{3}{2(1+x)^2} + C $$
$$ \int \frac{3}{(1+x)^3}dx = \left[-\frac{3}{2(1+x)^{2}}\right]_{0}^{c} $$
$$ = -\frac{3}{2(1+c)^{2}} - \left(-\frac{3}{2(1+0)^{2}}\right) $$
$$ = -\frac{3}{2(1+c)^{2}} + \frac{3}{2} $$
Now, Set the integral to 1:
$$-\frac{3}{2(1+c)^{2}} + \frac{3}{2} = 1 $$
$$-\frac{3}{2(1+c)^{2}} =  - \frac{1}{2} $$
$$ \frac{3}{2(1+c)^{2}} = \frac{1}{2} $$
$$ (1+c)^2 = 3  $$
$$ (1+c) = \sqrt{3}  $$
$$ c = \sqrt{3} - 1 = 0.732 $$
The value of c that makes f(x) a legitimate probability density function is: 0.732

\begin{itemize}
\item[b.] Use R and plot the pdf curve of the random variable.
\end{itemize}
```{r}
# PDF function
f <- function(x, c) {
  3 / (1 + x)^3
}

# Setting the value for C
c <- sqrt(3) - 1

# Generate x values for plotting
x <- seq(0, c, length.out = 100)

# Calculating the PDF values
y <- f(x, c)

# Plotting the PDF curve
plot(x, y, type = "l", col = "blue", xlab = "x", ylab = "f(x)", main = "PDF of the Random Variable")
```
\begin{itemize}
\item[c.] What is E(X)?
\end{itemize}
$$ E(X) = \int_{0}^{c} x*f(x)dx $$
$$ E(X) = \int_{0}^{c} x*\frac{3}{(1+x)^3}dx =  3\int_{0}^{c} \frac{x}{(1+x)^3}dx $$
Applying integration by parts:
$$ \int u\ dv = uv - \int v \ du $$
$$ \int \frac{x}{(1+x)^3}dx =  x \left( - \frac{1}{2(1+x)^2}\right) - \int -\frac{1}{2(1+x)^2}dx$$
$$ = -\frac{x}{2(1+x)^2} + \frac{1}{2}\int (1+x)^{-2} dx$$
Solving the integral:
$$ \int (1+x)^{-2} dx = - \frac{1}{1+x} $$
So, we have:
$$ \int \frac{x}{(1+x)^3}dx = -\frac{x}{2(1+x)^2} -\frac{1}{2(1+x)} + C$$
Evaluating with the limits [0,c]:
$$ \int \frac{x}{(1+x)^3}dx = \left[ -\frac{x}{2(1+x)^2} -\frac{1}{2(1+x)} \right]_{0}^{c}$$
At x=c:
$$ -\frac{c}{2(1+c)^2} -\frac{1}{2(1+c)} $$
At x=0:
$$ -\frac{0}{2(1+0)^2} -\frac{1}{2(1+0)} = - \frac{1}{2} $$
Therefore,
$$ 3\int \frac{x}{(1+x)^3}dx = 3\left( \left[ -\frac{c}{2(1+c)^2} -\frac{1}{2(1+c)} \right] - \left[- \frac{1}{2} \right] \right)$$
$$ \text{substituting } c=\sqrt{3}-1$$
$$ E(X) = 3\left( - \frac{\sqrt{3}-1}{2(\sqrt{3})^2} - \frac{1}{2\sqrt{3}} + \frac{1}{2} \right) = 0.268 $$
\begin{itemize}
\item[d.] Use R and simulate 1000 observations from this statistical population?
\end{itemize}
```{r}
# Setting the value for c
c <- sqrt(3) - 1

# Simulating 1000 observations
n <- 1000
x <- (1 + runif(1000, 0, c)) - 1

print(x)
```

\begin{itemize}
\item[e.] Use the generated data from part (d), estimate the mean and variance of the distribution?
\end{itemize}
```{r}

mean_x <- mean(x)

variance_x <- var(x)

print(paste("Estimated mean:", mean_x))
print(paste("Estimated variance:", variance_x))
```
***
## Question 6
\begin{itemize}
\item[1.] Write a Python function which takes a list of numbers x and returns a dictionary including
\end{itemize}
- Min : minimum value of x
- $Q_1$ : first quartile of x
- M : median of x
- $Q_3$ : third quartile of x
- Max : maximum value of x
- IQR : Q3 − Q1,
- Outliers : a list of x values which are either smaller than Q1−1.5×IQR or greater than Q3+1.5×IQR.
```{python}
def statistics(x):

  # Calculating the required statistics
  min_x = min(x)
  q1 = np.percentile(x, 25)
  median_x = np.median(x)
  q3 = np.percentile(x, 75)
  max_x = max(x)
  iqr = q3 - q1
  outliers = [value for value in x if value < q1 - 1.5 * iqr or value > q3 + 1.5 * iqr]


  stats_dict = {
    "Min": min_x,
    "Q_1": q1,
    "M": median_x,
    "Q_3": q3,
    "Max": max_x,
    "IQR": iqr,
    "Outliers": outliers
  }

  return stats_dict
```
\begin{itemize}
\item[2.] Apply your function in part (1) to the following list
\end{itemize}
$$ x = [2, 36, 12, 14, 204, 21.6, 22.5, 1, 32.8, 32.1, 13, 10, 88, 3.3, 3.1, 88] $$
```{python}
x = [2, 36, 12, 14, 204, 21.6, 22.5, 1, 32.8, 32.1, 13, 10, 88, 3.3, 3.1, 88]

stats_dict = statistics(x)

for key, value in stats_dict.items():
    print(f"{key}: {value}")
```
***
## Question 7
\begin{itemize}
\item[2.]
\end{itemize}
```{python}
def leaveOneOutCrossValidation(X, y):

  n = X.shape[0]  # Number of observations
  y_pred = np.zeros(n)  # Vector to store predictions

  # Iterating through each observation
  for i in range(n):
    # Removing the i-th observation from X and y
    X_train = np.delete(X, i, axis=0)
    y_train = np.delete(y, i)

    # Train the model using remaining observations
    beta_star = np.linalg.solve(X_train.T @ X_train, X_train.T @ y_train)

    # Predicting the response for removed observation
    y_pred[i] = X[i, :] @ beta_star

  # root mean squared error
  rmse = np.sqrt(np.sum((y - y_pred)**2) / n)

  return rmse
```
\begin{itemize}
\item[3.]Load the diabetes data from sklearn package in python and take a random sample of 56 observations from the data set and compute your X as the first three features of these 56 observations. The target of these observations will be your response vector.
\end{itemize}
```{python}
from sklearn.datasets import load_diabetes

diab = load_diabetes()

# Extract the features (X) and target (y)
X = diab.data[:, :3]  # Take the first three features
y = diab.target

# Randomly select 56 observations
n = 56
random_indices = np.random.choice(X.shape[0], size=n, replace=False)
X_selected = X[random_indices]
y_selected = y[random_indices]
```
\begin{itemize}
\item[4.]Apply your function form part 2 to the data set of part 3 and report the root MSE.
\end{itemize}
```{python}

rmse = leaveOneOutCrossValidation(X_selected, y_selected)

print("Root Mean Squared Error (RMSE):", rmse)
```








